Displacement of fishing effort by Large Scale Marine Protected Areas

### 2018-10-13 rmd/Manuscript.Rmd
Updated intro

---
title: "Displacement of fishing effort by Large Scale Marine Protected Areas"
subtitle: "Updated on `r Sys.Date()`"
author: |
  Juan Carlos Villaseñor-Derbez^1^
  John Lynham^2^
abstract: "Large-scale Marine Protected Areas (LSMPAs) have seen a significant increase over the last years. Fishing effort is effectively eliminated within these protected areas upon implementation. The benefits of reducing effort have been largely studied, and include increases in abundance, biomass, and diversity within the bounded regions. These no-take zones may produce spillover effects, which provide fish for outside areas. However, the economic and ecological implications of displacing fishing effort are not yet fully understood. Novel data products that track fishing effort at the vessel-level allow us to identify changes in fleet- and vessel-level behavior upon the implementation of protected areas, as well as how these redistribute. This papers evaluates the implications of implementing LSMPA, by evaluating changes in fishing hours, showing that vessels in the effected region reduce fishing effort after the implementation of PIPA. Our results are robust to a set of specifications. We also track the relative spatial allocation of fishing events thorugh time, and identify that areas closer to PIPA show an increase in relative fishing hourse due to the displacement of PIPA-fishing vessels. Our results not only provide an impact evaluation of the effect of LSMPAs on fishing activity, but provide insights into vessel redistribution dynamics, which may have ecological and economic implications."
date: "
^1^Bren School of Environmental Science and Management, University of California Santa Barbara, Santa Barbara, CA\\newline
^2^Department of Economics, University of Hawaii at Manoa, Honolulu, HI"
output:
  bookdown::pdf_document2:
    toc: no
    citation_package: natbib
    keep_tex: yes
    number_sections: no
fontsize: 12pt
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{lineno}
  - \linenumbers
bibliography: references.bib
# biblio-style: nature
---

@@

Marine Protected Areas (MPAs) are intended to safeguard parts of the ocean from extractive activities such as fishing. Current international goals aim to "effectively protect 10\% of the ocean environments by 2020". In an effort to meet this target, the world has seen a rapid increase in MPA coverage \citep{wood_2008,sala_2018b}. A significant part of this rapid increase can be attributed to the designation of a small number of Large Scale Marine Protected Areas [LSMPAs; \citep{gray_2017}]. These are defined as MPAS with at least 250,000 km^2^ in extension \citep{toonen_2013}, and are often implemented in the pelagic environment, where the dominant human activity is industrial fishing \citep{gray_2017}.

Due to weak property rights, limited habitat transformation, and potentially lower management costs, pelagic MPAs provide a great opportunity to safeguard the oceans \citet{game_2009}[^d]. A growing body of literature has also shown that closing the high seas to all fishing -effectively, turning them into a LSMPA- could increase fishery yields and profitability of fisheries, without negligible costs to food security \citep{white_2014,sumaila_2015,sala_2018,schiller_2018}. However, as with customary MPAs, it is important that we understand the socioeconomic implications of management interventions.

Given the relatively recent establishment of most LSMPAs, very little is known about their human dimensions and implication for fisheries \citep{gray_2017}. LSMPAs were erroneously assumed to have little social implications due to their remoteness \citep{agardy_2011,gray_2017}. However, the anticipation of a LSMPA can lead to preemptive overfishing, which can erode or delay the expected benefits of the intervention \citep{mcdermott_2018}. LSMPAs have received great attention in terms of governance and enforcement, but are yet to be the focus of economic analyses \citep{gray_2017}. For example, \citet{mcdermott_2018} show that fishing effort within the Phoenix Islands Marine Protected Area (PIPA) is effectively reduced after implementation, and describe changes in fishing behavior pre-implementation. \citet{cabral_2017} analyse the redistribution of fishing and non-fishing vessels following the implementation of MPAs in California, and find that responses are idiosyncratic; commercial dive boats follow a fishing-the-line pattern, while some fishing boats follow an ideal free distribution. The way in which fishers react to a spatial closure can have major implications in its outcome \citep{hilborn_2006,krueck_2017,viana_2017}. This highlights the need to understand how fishers react to the implementation of a LSMPA, and fishing effort changes and is spatially redistributed.

The main objective of this paper is to identify how fishers adapt to the implementation of LSMPAs. We combine novel vessel tracking technologies and causal inference techniques to identify behavioral changes of fishing vessels due to the implementation of PIPA. We focus on fishing hours and distance traveled as outcome variables that fishers might adjust following implementation of a LSMPA in an impact-evaluation fashion. Additionally, we evaluate the spatial redistribution of fishing effort that existed within PIPA before its implementation. This work provides novel empirical insights into fisher's responses to the implementation of LSMPAs, and can help guide future interventions.

@@

```{r}
effort_by_vessel <- readRDS(file = here::here("data", "effort_by_vessel.rds")) %>% 
  filter(year < 2018)
```

```{r}
# Just to get pipa vessels that fished inside pipa before 2015
pipa_vessels <- effort_by_vessel %>% 
  filter(treated, baci_strict) %$%
  unique(mmsi) %>% 
  length()
```

@@

# Discusion

@@

```{r}
models_exp0_did
models_exp1_did
models_exp2_did
models_exp3_did
```

>>>>

---
title: "Displacement of fishing effort by Large Scale Marine Protected Areas"
subtitle: "Updated on `r Sys.Date()`"
author: |
  Juan Carlos Villaseñor-Derbez^1^
  John Lynham^2^
abstract: "Large-scale Marine Protected Areas (LSMPAs) have seen a significant increase over the last years. Fishing effort is effectively eliminated within these protected areas upon implementation. The benefits of reducing effort have been largely studied, and include increases in abundance, biomass, and diversity within the bounded regions. These no-take zones may produce spillover effects, which provide fish for outside areas. However, the economic and ecological implications of displacing fishing effort are not yet fully understood. Novel data products that track fishing effort at the vessel-level allow us to identify changes in fleet- and vessel-level behavior upon the implementation of protected areas, as well as how these redistribute. This papers evaluates the implications of implementing LSMPA, by evaluating changes in fishing hours, showing that vessels in the effected region reduce fishing effort after the implementation of PIPA. Our results are robust to a set of specifications. We also track the relative spatial allocation of fishing events thorugh time, and identify that areas closer to PIPA show an increase in relative fishing hourse due to the displacement of PIPA-fishing vessels. Our results not only provide an impact evaluation of the effect of LSMPAs on fishing activity, but provide insights into vessel redistribution dynamics, which may have ecological and economic implications."
date: "
^1^Bren School of Environmental Science and Management, University of California Santa Barbara, Santa Barbara, CA\\newline
^2^Department of Economics, University of Hawaii at Manoa, Honolulu, HI"
output:
  bookdown::pdf_document2:
    toc: no
    citation_package: natbib
    keep_tex: yes
    number_sections: true
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{lineno}
  - \linenumbers
bibliography: references.bib
# biblio-style: nature
---

@@

Marine Protected Areas (MPAs) are intended to safeguard parts of the ocean from fishing and other extractive activities. Current international goals aim to "effectively protect 10\% of the ocean environments by 2020". In an effort to meet this target, the world has seen a rapid increase in MPA coverage \citep{wood_2008,sala_2018}. A significant part of this rapid increase can be attributed to the designation of a small number of Large Scale Marine Protected Areas [LSMPAs; \citep{gray_2017}]. These are defined as MPAS with at least 250,000 Km^2^ in extension (Fig. \ref{fig:LSMPAs_map}), and are often implemented in the pelagic environment, where the dominant human activity is industrial fishing \citep{toonen_2013,gray_2017,kroodsma_2018}.

Due to weak property rights, limited habitat transformation, and potentially lower management costs, pelagic MPAs provide an opportunity to safeguard the oceans \citet{game_2009}. A growing body of literature has shown that closing the high seas to all fishing -effectively, turning them into a LSMPA- could increase fishery yields and profitability of fisheries, with negligible costs to food security \citep{white_2014,sumaila_2015,sala_2018a,schiller_2018}. However, as with customary MPAs, it is important that we understand the socioeconomic implications of management interventions.

Given the relatively recent establishment of most LSMPAs, very little is known about their human dimensions and implication for fisheries \citep{gray_2017}. LSMPAs were erroneously assumed to have little social implications due to their remoteness \citep{agardy_2011,gray_2017}. However, the anticipation of a LSMPA can lead to preemptive overfishing, which can erode or delay the expected benefits of the intervention \citep{mcdermott_2018}. LSMPAs have received great attention in terms of governance and enforcement, but are yet to be the focus of economic analyses \citep{gray_2017}.

Research on LSMPAs have been the focus of analysing X, Y, and Z. However, no studies look into displacement of fishing effort. Redistribution of fishing effort has been evaluated in ssf by X, Y ,and Z. \citet{cabral_2017} analyse the redistribution of fishing and non-fishing vessels following the implementation of MPAs in California, and find that responses are idiosyncratic; commercial dive boats follow a fishing-the-line pattern, while some fishing boats follow an ideal free distribution. The way in which fishers react to a spatial closure can have major implications in its outcome \citep{smith_2003,hilborn_2006}. This highlights the need to understand how fishers react to the implementation of a LSMPA, and fishing effort changes and is spatially redistributed.

The main objective of this paper is to identify how fishers change their behavior due to the implementation of LSMPAs. We combine novel vessel tracking technologies and causal inference techniques to identify behavioral changes of fishing vessels due to the implementation of PIPA. We focus on fishing hours and distance traveled as outcome variables that fishers might adjust following implementation of a LSMPA in an impact-evaluation fashion. Additionally, we evaluate the spatial redistribution of fishing effort that existed within PIPA before its implementation. This work provides novel empirical insights into fisher's responses to the implementation of LSMPAs, and can help guide future interventions. Our work is novel in the sense that it provides empirical evidence of the effect of large-acale Marine Protected Areas in fishing behavior and distribution.

The next sections are as follows: Section \ref{background} provides an overview of the Nauru Agreement and associated countries, a description of the fleet that operates in the region, and a brief history of PIPA. Section \ref{methods} describes our data and identification strategy. Section \ref{results} presents our results, section \ref{discussion} provides an extension of our results to other LSMPAs and discusses our results.

```{r, fig.cap = "\\label{fig:LSMPAs_map}"}

mpas <- read_sf(dsn = here::here("data", "spatial", "LSMPAs"), layer = "LSMPAs") %>% 
  st_rotate() %>% 
  mutate(PIPA = wdpaid == 309888)

coast <- map_data('world', wrap=c(-20,340), ylim=c(-85,75))%>%
  select(group, lat, long) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
  group_by(group) %>% 
  summarize(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON") %>% 
  mutate(a = 1) %>% 
  group_by(a) %>% 
  summarize() %>% 
  ungroup()

pipa <- sf::read_sf(dsn = here::here("data", "spatial", "PIPA"), layer = "PIPA") %>% 
  sf::st_transform(crs = "+proj=longlat +datum=WGS84 +no_defs") %>% 
  st_rotate()

ggplot() +
  geom_sf(data = coast) +
  geom_sf(data = mpas, fill = "transparent", color = "black", size = 1) +
  geom_sf(data = mpas, aes(fill = PIPA), color = "transparent") +
  ggtheme_map() +
  scale_fill_brewer(palette = "Set1", direction = -1) +
  theme(legend.position = "none")
```

# Background

The Nauru Agreement was established in 1982 by Pacific island nations to manage their important tuna resources. PNA Members include Federated States of Micronesia, Kiribati, Marshall Islands, Nauru, Palau, Papua New Guinea, Solomon Islands, and Tuvalu.The Nauru Agreement regulated access of foreign vessels (*i.e.* those from non-PNA countries). Holding ~80% of the historical purse seining grounds within their Exclusive Economic Zones, PNA countries gained bargaining power when providing access to foreign fleets \citep{havice_2010}.

The cooperation that emerged thanks to the PNA allowed for subsequent agreements that strengthened fisheries management, like the Palau Agreement, which limited the number of purse seiners at 205 vessels from 1995-2007[^a]. However, the most notable regulation is their approach to manage fishing effort: a Vessel Day Scheme (VDS) implemented in 2007 \citep{havice_2013}. This effectively modified how fishing effort was managed, from number of vessels under the Palau Agreement to fishing hours. The VDS works as follows: Each year, scientific advisors recomend a total number of fishing vessel-days per year. Hours are allocated to each PNA country based on catch history, and they then sell fishing rights to other non-PNA countries \citep{aqorau_2018}. While the effectiveness of this scheme has been debated in terms of meeting their fishery management and conservation objectives, the licensing significantly contributes to the economy of these island nations \citep{havice_2010}.

[^a]: See \citet{havice_2010} for a detailed description of the Nauru, Palau, and Federal States of Micronesia agreements, their objectives, and outcomes.

```{r, fig.cap = "\\label{fig:PNA_map}"}
countries <- c("PIPA", "KIR", "HS", "ASM", "COK", "FSM", "MHL", "NRU", "PNG", "SLB", "TKL", "TUV", "UMI")

small_coast <- rnaturalearth::ne_countries(returnclass = "sf") %>% 
  filter(sov_a3 %in% countries) %>% 
  st_rotate()

eez <- read_sf(dsn = here::here("data", "spatial", "EEZ_subset"),
               layer = "EEZ_subset") %>% 
  filter(ISO_Ter1 %in% countries) %>% 
  mutate(KIR = ISO_Ter1 == "KIR") %>% 
  group_by(ISO_Ter1, PNA, KIR) %>% 
  summarize()

ggplot() +
  geom_sf(data = small_coast) +
  geom_sf(data = eez, aes(color = KIR, linetype = !PNA), fill = "transparent") +
  geom_sf(data = pipa, fill = "red") +
  ggtheme_plot() +
  scale_color_manual(values = c("black", "red")) +
  theme(legend.position = "none")
```

The main tuna species caught in the region are skipjack (*Katsuwonus pelamis*), yellowfin (*Thunnus albacares*), albacore (*Thunnus alalunga*) and bigeye (*Thunnus obesus*). From these, the first two are amongst the top-10 species represented in global fisheries production statistics, with catches increasing relative to the 2005-2014 average \citep{fao_2018}. The pacific region has historically accounted for a large portion of tuna species production \citep{aqorau_1997}. Today, the PNA controls close to 50% of the global skipjack tuna production \citep{pna_website_2018}. A large portion of these catches derive from purse seine vessels licensed under the VDS. Fishing vessels from Australia, New Zeland, China, France, Korea, Japan, the Philippines, Taiwan and the United States participate in the purse-seining vessel day schemes.

For example, \citet{mcdermott_2018} show that fishing effort within the Phoenix Islands Marine Protected Area (PIPA) is effectively reduced after implementation, and describe changes in fishing behavior pre-implementation.

@@

```{r, eval = T}
effort_by_vessel <- readRDS(file = here::here("data", "effort_by_vessel.rds")) %>% 
  filter(year < 2018)
```

```{r, eval = T}
# Just to get pipa vessels that fished inside pipa before 2015
pipa_vessels <- effort_by_vessel %>% 
  filter(treated, baci_strict) %$%
  unique(mmsi) %>% 
  length()
```

@@

# Discussion

@@

```{r}
models_exp0_did %>% stargazer::stargazer()
models_exp1_did %>% stargazer::stargazer()
models_exp2_did %>% stargazer::stargazer()
models_exp3_did %>% stargazer::stargazer()
```
